---
title: "Final Project (Question 9 & 10)"
author: "Muhammad Idris"
date: "11/18/2020"
output: powerpoint_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load data and libraries
```{r}
library(ggplot2)
library(tidyverse)
library(ggrepel)
library(gganimate)
library(gifski)
#install.packages('png')
```

```{r}
athlete_data <- read.csv('athlete_events.csv')
regions <- read.csv('noc_regions.csv')
```

```{r}
athlete_data$Sex <- athlete_data$Sex %>% as.factor()
#athlete_data$Season <- as.factor()
athlete_data$Height <- as.double(athlete_data$Height)
athlete_data$Weight <- as.double(athlete_data$Weight)
athlete_data$Medal <- athlete_data$Medal %>% factor(levels = c('Gold','Silver','Bronze'))
```

```{r}
data.frame(athlete_data)
```  

```{r}
# Total missing values
sum(is.na(athlete_data))

# Missing values by column
colSums(is.na(athlete_data))
```

## Missing Values

- A missing value in the Medals category could indicate that the athlete did not win a medal for that particular Event, Year, and Games.


```{r}
athlete_cleaned <- athlete_data

medal_levels <- levels(athlete_cleaned$Medal)

medal_levels[length(medal_levels) + 1] <- "None"

athlete_cleaned$Medal <- factor(athlete_cleaned$Medal, levels = medal_levels)

athlete_cleaned$Medal[is.na(athlete_cleaned$Medal)] <- "None"

data.frame(athlete_cleaned)

```



```{r}
length(unique(athlete_cleaned$Team))
``` 


```{r}
athlete_region <- athlete_cleaned %>% left_join(regions,by="NOC") %>% filter(!is.na(region)) 

data.frame(athlete_region)
```


## 9. Which individual has most number of medals (Team = Individual Olympic Athletes )

- Athletes have competed as Independent Olympians at the Olympic Games for various reasons, including political transition, international sanctions, suspensions of National Olympic Committees.

```{r ,echo = TRUE}
ioa <- athlete_region %>% filter(Team == 'Individual Olympic Athletes') 

ioa
```



```{r, echo = TRUE}
#Number of medals awarded to Individual Olympic Athletes from 1896 to 2016
ind_medal <- athlete_region %>% filter(Team == 'Individual Olympic Athletes', Medal != 'None') %>% group_by(Name, Sport, Medal) %>% summarize(Count=length(Medal)) 

# order Athlete by total medal count
medal_count <- ind_medal %>% group_by(Name) %>% summarize(Total=sum(Count)) %>% arrange(Total) %>% select(Name)

ind_medal$Name <- factor(ind_medal$Name, levels=medal_count$Name)

#ind_medal <- ind_medal %>% arrange(desc(Count))

ind_medal
```  

```{r}
ind_medal %>% ggplot(aes(x = Name, y = Count, fill = Medal)) + geom_col() + xlab('Athlete') + ylab('Count') + coord_flip() + ggtitle('Individual Olympic Athletes Medal Count') + theme(plot.title = element_text(hjust = 0.5)) 
```  

## 10.What Olympic events have stayed around, which have been removed?

```{r}
# Count Events each year
counts_event <- athlete_cleaned %>% filter(Team != "Unknown") %>% group_by(Year,Season) %>%
  summarize(Events = length(unique(Event)))

counts_event
```

```{r pressure}
# # Change of events over time
# ggplot(counts_event, aes(x=Year, y=Events, group = Season, color = Season)) + geom_point(size=1) + geom_line() + xlab("Olympic Events from 1896 to 2016")

animated <- ggplot(counts_event, aes(x=Year, y=Events, group=Season, color=Season)) + geom_point(size=2) + geom_line()  + transition_reveal(Year) + labs(title = "Olympic Events from 1896 to 2016") + theme(plot.title = element_text(hjust = 0.5))

animate(animated)
# anim_save("events.gif")
```  

- Highest number of events in 1992 with 314 events. 

- Filtered Year before 2000 and after 2000

-Used anti_join and left_join to find missing events 

https://www.topendsports.com/events/discontinued/list.htm#:~:text=Since%20the%20first%20modern%20Games,%2C%20softball%2C%20and%20motor%20boating.

```{r}
before2000 <- athlete_region %>% filter(Year >= 1896 & Year< 2000) %>% select(Event, Year)

before2000 
```
```{r}
length(unique(before2000$Event))
```



```{r}
after2000 <- athlete_region %>% filter(Year >= 2000) %>% select(Event, Year) 

after2000
```  

```{r}
length(unique(after2000$Event))
```



```{r}
discont_events <- anti_join(before2000, after2000, by= "Event")

discont_events

```
```{r, echo = TRUE}
unique(discont_events$Event)
```  

```{r}
discont_events2 <- anti_join(after2000, before2000, by= "Event")

discont_events2
```  

```{r, echo = TRUE}
joined2 <- left_join(discont_events, discont_events2, by = 'Event')

joined2

unique(joined2$Event)
```   
```{r}
joined <- left_join(discont_events2, discont_events, by = 'Event')

joined

unique(joined$Event)
```


-Some of the discontinued events include Art Competitions (1912-1948),Tug-Of-War Men's Tug-Of-War(1900-1920),Military Ski Patrol Men's Military Ski Patrol.

```{r}

#summary(comparedf(discont_events, discont_events2, by = 'Event'))

comparison <- semi_join(after2000, before2000, by ='Event')

comparison

length(unique(comparison$Event))
```



