---
title: "Patrick"
author: "Patrick Origer"
date: "11/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
library(tidyverse)
library(dplyr)
library(tidyr)
library(stringr)
library(lubridate)
```

```{r}
olympics <- read.csv('athlete_events.csv', stringsAsFactors = FALSE)
```

-Line chart displaying number of athletes through time by season
```{r}
olympics %>% group_by(Year, Season) %>%
  summarise(NumberOfAthletes = n_distinct(Name)) %>%
  ggplot(aes(x = Year, y = NumberOfAthletes, color = Season)) +
  geom_line() + geom_point() +
  labs(title = "Number of Athletes by Season",
       y = "Number of Athletes")
```

-Creating data frame in regards to a gender gap
```{r}
sexGap <- olympics %>%
  group_by(Sex, Year) %>%
  summarise(NumAthletes = n_distinct(Name)) %>%
  pivot_wider(names_from = Sex, values_from = NumAthletes) %>%
  rename(NumberFemaleAthletes = F, NumberMaleAthletes = M) 

sexGap[is.na(sexGap)] = 0

sexGap <- sexGap %>%
  mutate(FemaleToMaleRatio = NumberFemaleAthletes / NumberMaleAthletes, 
  FemaleAndMaleDifference = NumberMaleAthletes - NumberFemaleAthletes)
```

-Table displaying ratios of sex by year
```{r}
sexGapSummary <- sexGap %>% select(Year, FemaleToMaleRatio) %>%
  mutate(MaleToFemaleRatio = 1 / FemaleToMaleRatio) %>%
  arrange(MaleToFemaleRatio)

head(sexGapSummary, 10)
```

3. How has the gender gap changed over time?
The gender gap has drastically shrunk from its beginning to now.

-Line plots displaying displaying ratio and difference of athletes by sex
```{r}
ggplot(sexGap, aes(x = Year, y = FemaleToMaleRatio)) + geom_line() + geom_point() +
  labs(title = "Ratio Comparison From Female to Male Athletes", y = "Female to Male Ratio")

ggplot(sexGap, aes(x = Year, y = FemaleAndMaleDifference)) + geom_line() + geom_point() +
  labs(title = "Difference in Number of Athletes Between Sexes", y = "Sex Difference")
```

-Creating two data frames in regards to their season
```{r}
sexGapWinter <- olympics %>%
  filter(Season == "Winter") %>%
  group_by(Sex, Year) %>%
  summarise(NumAthletes = n()) %>%
  pivot_wider(names_from = Sex, values_from = NumAthletes) %>%
  rename(NumberFemaleAthletes = F, NumberMaleAthletes = M) %>%
  mutate(FemaleToMaleRatio = NumberFemaleAthletes / NumberMaleAthletes, 
  FemaleAndMaleDifference = NumberMaleAthletes - NumberFemaleAthletes) %>%
  arrange(desc(FemaleToMaleRatio))

sexGapSummer <- olympics %>%
  filter(Season == "Summer") %>%
  group_by(Sex, Year) %>%
  summarise(NumAthletes = n()) %>%
  pivot_wider(names_from = Sex, values_from = NumAthletes) %>%
  rename(NumberFemaleAthletes = F, NumberMaleAthletes = M) 

sexGapSummer[is.na(sexGapSummer)] = 0

sexGapSummer <- sexGapSummer %>%
  mutate(FemaleToMaleRatio = NumberFemaleAthletes / NumberMaleAthletes, 
  FemaleAndMaleDifference = NumberMaleAthletes - NumberFemaleAthletes) %>%
  arrange(desc(FemaleToMaleRatio))

```

-Line plot displaying female to male athlete ratio
```{r}
ggplot(sexGapWinter, aes(x = Year, y = FemaleToMaleRatio)) +
  geom_line(aes(color = "Winter")) + geom_point(aes(color = "Winter")) +
  geom_line(data = sexGapSummer, aes(color = "Summer")) + geom_point(data = sexGapSummer, aes(color = "Summer")) +
  scale_color_manual(name = "Season", values = c("red", "blue")) +
  labs(title = "Ratio of Athletes by Sex Between Type of Season",
       y = "Female to Male Athlete Ratio")
```

-Line plot displaying female and male difference in number of athletes
```{r}
ggplot(sexGapWinter, aes(x = Year, y = FemaleAndMaleDifference)) +
  geom_line(aes(color = "Winter")) + geom_point(aes(color = "Winter")) +
  geom_line(data = sexGapSummer, aes(color = "Summer")) + geom_point(data = sexGapSummer, aes(color = "Summer")) +
  scale_color_manual(name = "Season", values = c("red", "blue")) +
  labs(title = "Difference in Number of Athletes Between Sexes by Season",
       y = "Male and Female Athlete Difference")
```

4. Are there any missing years, if so, why?
There have been multiple times when the Olympics were not been held.
  -1916 Olympics were planned to be held in Berlin, Germany, but that was canceled due to the outbreak of World War I.
  -1940 Olympic games, in both winter and summer, were canceled due to World war II.
  -1944 Olympic games, in both winter and summer, were canceled due to World war II.
  -Winter Olympics began occurring two years apart from the Summer Olympics every four years in 1994.
  
```{r}
ggplot(sexGap, aes(x = Year, y = FemaleAndMaleDifference)) + geom_bar(stat = "identity", fill = "skyblue4") + 
  labs(title = "Difference in Number of Athletes Between Sexes", y = "Male and Female Athlete Difference")

sexGap %>% filter(Year >= 1904 & Year <= 1924) %>% ggplot(aes(x = Year, y = FemaleAndMaleDifference)) +
  geom_bar(stat = "identity", fill = "skyblue4") +
  labs(title = "Difference in Number of Athletes Between Sexes (1904-1924)", y = "Male and Female Athlete Difference") +
  geom_text(aes(label = Year), vjust = -0.25)

sexGap %>% filter(Year >= 1932 & Year <= 1952) %>% ggplot(aes(x = Year, y = FemaleAndMaleDifference)) +
  geom_bar(stat = "identity", fill = "skyblue4") +
  labs(title = "Difference in Number of Athletes Between Sexes (1932-1952)", y = "Male and Female Athlete Difference") +
  geom_text(aes(label = Year), vjust = -0.25)

sexGap %>% filter(Year >= 1984 & Year <= 2000) %>% ggplot(aes(x = Year, y = FemaleAndMaleDifference)) +
  geom_bar(stat = "identity", fill = "skyblue4") +
  labs(title = "Difference in Number of Athletes Between Sexes (1984-2000)", 
       y = "Male and Female Athlete Difference") +
  geom_text(aes(label = Year), vjust = -0.25)
```










